An Experimental Analysis of the Influence of the Precision of Parameters in an Artificial Neural Network

### 2025-07-16 paper/main.tex
added decimal roudning content

\subsection*{Effect of Decimal Rounding}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/118_accuracy.png}
	\caption{Accuracy vs.\ number of decimal digits retained in weights.}
\end{figure}

>>>>

\subsection*{Effect of Decimal Rounding}

A non training related issue is the storage of the NN parameters. Usually, the parameters
are stored in the format that was chosen to train with. This is best for reproducibility
as well as accuracy when doing inference with the trained model weights. The datatype  
precision has a higher impact during training though, which begs the question if one could
omit precision from the model weights. In this section, we explore a naive quantization
method: Rounding to a fixed number of decimals. With significantly little decimals and a
reasonable range for the integer part, one could store only the significant and provide
base as well as exponent as metadata.

The first step is to examine the distribution of parameter values. By the standard
deviation around modes one can make an educated guess as to how many decimals are needed.
The range will decide how large the integer part may get. Reading from the 
distribution is just a heuristic though, so we will also test the hypotheses.

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/118_dist.png}
	\caption{
		Distribution of model weights in each layer. The values are in $[-0.5, 0.5]$,
		this means no integer part is needed in this example. Generally, with normalized data
		and the quite popular L1-/L2-/Batch-Norm layers, weight values in $[-1, 1]$ is to be expected.}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/118_dist2.png}
	\caption{Distribution of model weights in each layer after rounding to two decimals.
	This appears to be a good approximation already.}
	\label{fig:118_2}
\end{figure}

\begin{figure}[H]
	\centering
	\includegraphics[width=0.85\textwidth]{figures/118_accuracy.png}
	\caption{Test Accuracy vs.\ number of decimal digits retained in weights.}
	\label{fig:118_acc}
\end{figure}

Figure~\ref{fig:118_acc} shows how rounding the decimal part affects the test accuracy.
For just one decimal the model has a maximum of 19 different weights at its disposal,
which makes the model have only 10\% accuracy, random guessing with a 10 class problem.
Two decimals perform significantly better. At 70\% accuracy the model gets close to the
full precision 88\% accuracy with 32 bit floating point. Making and educated guess on
distribution such as Figure~\ref{fig:118_2} seems to be a good heuristic for this example.
At just 3 decimals, the full 32 bit precision is reached. Thus we can conclude that with
just 10 bit for representing 3 decimals, no bits for the integer part and one additional
bit for the sign, only 11 bit must be stored to have the best performing model for
inference.
